---
title: "PS_5"
format: html
editor: visual
---

## **Problem 1 - Plotting**

Use the \"nnmaps\" data set again. I\'d recommend using **ggplot2** for these, but you can do them in base R or **plotly** if you\'d prefer.

```{r}
library(assertthat)
library(ggplot2)
nnmaps <- read.csv("/Users/dereklief/Desktop/MAIN FOLDER/Classes and Other/STATS 506/PS 2/stats506_ps5/chicago-nmmaps.csv")
```

a.  Produce a graph of the **mean** monthly temperature in **celsius**. Draw a scatter-plot where the x-axis is month and the y-axis is the average monthly temperature in celsius. Add a line connecting the points within each season and color the lines and points by season (same color for lines and points in the same season). Be sure both x-axis and the legend are ordered appropriately, and that all labels/titles are clear and ready for publication (e.g. no variable names).

```{r}
library(dplyr)
library(ggplot2)
library(scales)
df <- nnmaps
df$temp_cels <- (df$temp - 32) * (5/9)

df_2 <- df %>% mutate(date = as.POSIXct(date, format = '%Y-%m-%d'), year_2 = format(date, "%Y"), yearmon = format(date, "%Y-%m")) %>%
  group_by(year, yearmon, month, season) %>%
  summarize(mean_cels = mean(temp_cels))
df_2$yearmon_2 <- paste0(df_2$yearmon, '-01')
df_2$yearmon_2 <- as.Date(df_2$yearmon_2, format = '%Y-%m-%d')
df_2$yearmon <- (paste("01-","",df_2$yearmon))

ggplot(df_2, aes(x = yearmon_2, y = mean_cels, color = season, group = season)) +
  geom_point() +
  geom_line() +
  ggtitle("Mean Monthly Temperature in Celsius") +
  scale_x_discrete(name = "Month-Year") +
  scale_color_discrete(name = "Seasons") +
  scale_x_date(date_breaks = '1 month', date_labels = "%b-%Y") +
  xlab("Month-Year") +
  ylab("Temperature in Celsius") +
  theme(plot.title = element_text(color="black", size=12, face="bold"),
        axis.title.x = element_text(color="black", size=10, face="plain"),
        axis.title.y = element_text(color="black", size=10, face="plain"),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size = 6, angle = 90)) +
  scale_color_manual("Seasons",values=c("brown", "green4", "gold3", "blue4")) 
```

b\. Produce a similar plot with four sets of lines: mean monthly temperature in celsius, mean montly O3, mean monthly PM10, and mean monthly dewpoint. Figure out a way to distinguish between the four lines. Again, make sure the plot is publication-ready (clean and clear labels). Use the plot to answer the following question:

```{r}
df_3 <- df %>% mutate(date = as.POSIXct(date, format = '%Y-%m-%d'), year_2 = format(date, "%Y"), yearmon = format(date, "%Y-%m")) %>%
  group_by(year, yearmon, month, season) %>%
  summarize(mean_cels = mean(temp_cels, na.rm = TRUE),
            mean_o3 = mean(o3, na.rm = TRUE),
            mean_pm10 = mean(pm10, na.rm = TRUE),
            mean_dp = mean(dewpoint, na.rm = TRUE))

df_3$yearmon_3 <- paste0(df_3$yearmon, '-01')
df_3$yearmon_3 <- as.Date(df_3$yearmon_3, format = '%Y-%m-%d')

ggplot() +
  geom_line(df_3, mapping = aes(x=yearmon_3, y = mean_cels, color = "red4")) +
  geom_point(df_3, mapping = aes(x=yearmon_3, y = mean_cels, color = "red4")) +
  geom_line(df_3, mapping = aes(x=yearmon_3, y = mean_o3, color = "blue4")) +
  geom_point(df_3, mapping = aes(x=yearmon_3, y = mean_o3, color = "blue4")) +
  geom_line(df_3, mapping = aes(x=yearmon_3, y = mean_pm10, color = "green4")) +
  geom_point(df_3, mapping = aes(x=yearmon_3, y = mean_pm10, color = "green4")) +
  geom_line(df_3, mapping = aes(x=yearmon_3, y = mean_dp, color = "orange2")) +
  geom_point(df_3, mapping = aes(x=yearmon_3, y = mean_dp, color = "orange2")) +
  facet_wrap(~season) +
  scale_x_date(date_breaks = '2 months', date_labels = "%b-%Y") +
  ggtitle("Monthly Means of Temperature, O3, PM10 and Dew Point") +
  xlab("Month-Year") +
  ylab("Averages") +
  theme(plot.title = element_text(color="black", size=12, face="bold"),
        axis.title.x = element_text(color="black", size=10, face="plain"),
        axis.title.y = element_text(color="black", size=10, face="plain"),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size = 6, angle = 90)) +
  scale_color_identity(name = "Type",
                       labels = c("Temp in Cels","O3","PM10","Dew Point"),
                       guide = "legend")
```

**\"Which of the four variables seems to have the least seasonal trend?\"**

We can see from the graph that "Temperature in Celsius" appears to have the least seasonal trend.

## **Problem 2 - OOP Programming**
